\hspace{1cm} Do syntactic and semantic similarity lead to interference effects?\newline Evidence from self-paced reading and event-related potentials using German

### 2023-10-16 paper/main.tex
fixed typo
changed all \cite to \textcite commands

% Useful packages
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
%\usepackage[colorlinks=true, allcolors=blue]{hyperref}
\usepackage[natbib,maxcitenames=2,bibstyle=authoryear, style=apa]{biblatex}
\usepackage{csquotes}
\DeclareLanguageMapping{american}{american-apa}
\bibliography{interferencebib}

@@

By contrast, syntactic interference is expected to be triggered if the distractor matches the \{+grammatical subject\} cue. This situation can occur when the relative clause  in \ref{ex:1_vandyke07} that modifies \textit{the resident} were to be changed to \textit{who said that the warehouse was dangerous}. Now, the distractor matches the \{+grammatical subject\}.

@@

Previous work has shown clear evidence for this kind of retrieval interference \parencite{vandyke07, vandyke_mcelree06, vandyke_mcelree2011, nicenboim,  vandyke_lewis03,jaeger_etal_2017}.
However, two important questions remain unanswered. First, as discussed in \cite{jaeger_etal_2017}, almost all the published studies on interference effects are severely underpowered. Low power has the consequence that published estimates will tend to be overestimates or misestimates of the effect size \parencite{vasishth2018_signficancefilter}. Over-/misestimates have important implications for theory evaluation. One example is discussed in \cite{vasishth2018_signficancefilter}, where a key prediction of surprisal theory \parencite{levy&keller_2013} could not be validated when a larger-sample study was conducted; by contrast, the original small-sample study  \parencite{levy&keller_2013} showed suggestive evidence consistent with the theory. Another example is the recent heated debate on prediction in sentence comprehension, which was largely focused on the results of low-powered studies, leading to potentially misleading conclusions \parencite{nicenboim_etal_2020}.
Second, although it is clear and uncontroversial  \parencite[e.g.,][]{vandyke07,mertzen} that syntactic and semantic cues are involved in sentence processing, it is not obvious which cues are relevant for a particular retrieval event. For example, \cite{dillon2013} argued, following \cite{Sturt2003}, that the gender cue was either downweighted or not used at all when an antecedent-reflexive dependency needs to be built; instead, the antecedent of a reflexive could be identified using the syntactic principle A of the binding theory, which correctly identifies the target for retrieval.  Using computational modeling, \cite{dillon2013} suggested that antecedent-reflexive dependencies were immune to interference, and that the reason for this is that principle A of the binding theory is the primary driver of the cue-based retrieval process \parencite[cf. ][]{jaeger_etal_2020,yadav2021individual}. This was a surprising discovery because a priori, cue-based retrieval accounts do not allow for exceptions to retrieval interference in particular syntactic dependency types.
More broadly, an important implication of the work initiated by \cite{dillon2013}, among others, is that  the precise details of which cues are used during retrieval can affect processing difficulty in possibly surprising ways.

@@

All experiments used the 2 $\times$ 2 design developed by \citet{vandyke07}; the design crosses syntactic and semantic interference.

@@

Syntactic interference is manipulated via the syntactic status of the distractor, which is either subject (\ref{ex:hisyn}) of an embedded clause, or part of a prepositional phrase (\ref{ex:losyn}). Following \cite{vandyke07}, the assumption here is that  during the retrieval of the subject, a distractor in any subject position causes more syntactic interference than a distractor inside a prepositional phrase.

@@

Indeed, previous work using the same design as in the present paper has also found semantic interference effects at the pre-critical region: both \cite{vandyke07} and \cite{mertzen} found such effects. \citeauthor{mertzen}'s (\citeyear{mertzen}) Figures 5 and 6 indicated reading time differences in earlier regions, especially for their German data (see their Figure 6). \cite{vandyke07} did not report reading times for the whole sentences / distractors;  \citeauthor{vandyke07} attributed the effects observed at the pre-critical region to plausibility differences between conditions, but as \cite{mertzen} also points out, encoding interference could be an explanation even in that study. Given these earlier findings, our results are consistent with the encoding interference explanation.

\cite{mertzen} observed both syntactic and semantic interference effects in the pre-critical region. 
They discuss several alternative explanations for their observed effects. In addition to encoding interference, \cite{mertzen} propose three other alternative explanations for their data: parafoveal-on-foveal effects, sentence structure confounds across conditions, and predictive processing effects. Our results cannot be explained by the parafoveal-on-foveal explanation, because there is no parafoveal preview in self-paced reading. The observed difference between the high and low semantic interference conditions is independent of the syntactic manipulation. Therefore, the sentence structure confound is also not a good explanation for the effect in our study. Regarding the predictive-processing explanation, \citet{mertzen} argued that the pre-critical adverb must attach to the upcoming the verbal phrase; this leads to an anticipatory creation of a verb phrase chunk in memory, which triggers a retrieval of the subject already at the pre-critical region. However, in our study, the effect started at the distractor and became smaller as the two pre-critical adverbs were read (see Figure \ref{fig:whole_sentence} A and B). The predictive-processing explanation would incorrectly predict that the effect begins at the second adverb, which immediately precedes the verb. Consequently, the prediction explanation is also ruled out in our data, rendering encoding interference as the most likely explanation of the effects found prior to the critical verb in our data. Crucially this line of reasoning should not be limited to the explanation of the pre-critical effects, but should be extended to the effects in the critical region as well. The reading times for high vs.\ low semantic interference conditions started to differ at the distractor and differed throughout the rest of the sentence. It is not possible to pinpoint whether the effect in the critical region is due to a retrieval initiated at this region, or still due to encoding and maintaining the distractor in memory, or a combination of the two kinds of interference.

\subsection{Comparison with previous findings}
The results of the present study stand in contrast to previous findings \citep{vandyke07, mertzen}. The present study is the only one to find semantic and no syntactic interference effect. The difference in observed effects could be explained by differences in statistical power or by methodological differences. The present study has higher statistical power than the previous studies. \citet{vandyke07} had 35 - 40 participants and 36 - 48 items in each of her three experiments, respectively. \cite{mertzen} tested 61 English speakers and 121 German speakers with 40 items each. In contrast, we tested 774 participants with at least 60 items each (a subset of 160 participants read all 120 items). Therefore, the present study is the study with the highest power to date using this 2 $\times$ 2 interference design. Consequently, the estimates from the present study are the most precise ones so far for this design. The smaller effect size in the present study is a consequence of higher power, and the magnitude of the effects observed is comparable to that of other large-scale reading studies \citep{nicenboim} and meta-analyses on interference \citep{jaeger_etal_2017}.

@@

Models were run with four chains and 20,000 iterations of which 2,000 were used as warm-up phase \citep{schad_etal_2022_BF}. For the calculation of Bayes factors and the corresponding sensitivity analysis, we defined a range of priors on the parameters of interest, assuming a range of effect sizes \citep{nicenboim_stats}. We chose slope priors which assume a wide range of effect sizes from $-0.2$ to $-11$\,$\mu V$. However, the more realistic a priori effect sizes probably range from $-0.2$ to at most $-2$\,$\mu V$; this is based on our expectation that the interference effect investigated here, being a subtle manipulation, is probably smaller in magnitude than the well-established  effect of cloze probability on the N400 amplitude analyzed by \cite{nicenboim_stats}, which ranges from $0$ to $5$\,$\mu V$ \parencite[see section 5.2.5][]{nicenboim_stats}.

@@

Consistent with the self-paced reading study, there is clear evidence for semantic interference, and evidence against syntactic interference.  How do these observed patterns compare to the theoretical predictions of cue-based retrieval theory? We turn to this point next by discussing the quantiative predictions of the \cite{Lewis2005} model.

%The to date rather small body of ERP studies on interference effects during sentence comprehension gives an indeterminate picture of the nature of interference effects in ERP measures. The present study adds a large data point to the body of literature which reported negativities or more specifically N400 effects for high vs.\ low interference conditions \citep{vasishth_drenhaus_2011, martinetal2014, schoknecht2022, lee_garnsey}. Therefore, it contributes to the debate of how interference effects manifest themselves electrophysiologically (interference effects have also been reported in differences in the P600, \cite{Tanner_etal_2017, martinetal2014}).

@@

 \cite{mertzen} presented simulations that show approximately equally sized interference effects 
for both the syntactic and semantic manipulations. Their model assumed that three cues were used at the verb:  \{$\pm$ subject, $\pm$ same clause, $\pm$ animacy\}. The upper part of Figure \ref{fig:model_predictions} reproduces the \cite{mertzen}   simulations. These are prior predictive distributions of the effects of interest, by sampling from a prior distribution ($Beta(a=4,b=6)$) for the free scaling parameter F and then generating data from the model while holding all other parameters in the model at fixed values. The F parameter scales the activation to a retrieval time. The prior $Beta(a=4,b=6)$ was chosen in earlier work \parencite{jaeger_etal_2020}.

@@

The absence of (and evidence against) syntactic interference is surprising because the psycholinguistic literature has clearly established that syntactic constraints play a role in sentence processing. For example, \cite{dillon2013} and \cite{Sturt2003} argued that principle A of the binding theory controls the retrieval process in the antecedent-reflexive dependency, effectively eliminating or attenuating interference from other overlapping features (such as number marking) of intervening nouns that cannot be an antecedent for the reflexive. For further evidence for this idea from individual differences, see \cite{yadav2021individual}. Even more relevant to the present study is the fact that \cite{mertzen} did find syntactic interference effects.

Could the absence of syntactic interference be reconciled with a cue-based retrieval account? We speculated that one possibility might be that the parser tracks hierarchical structure and that this hierarchical information is used when searching for a noun in memory. Specifically, the relevant syntactic cues that might be used at the verb is not  \{$\pm$ subject, $\pm$ same clause\} but rather a composite cue  \{$\pm$ subject-in-same-clause\} . In other words, in the spirit of the  \cite{dillon2013} and \cite{Sturt2003} proposal, that the parser intelligently targets only the syntactically relevant noun during antecedent retrieval, we evaluated the predictions of the model assuming that the cues are   \{$\pm$ subject-in-same-clause, $\pm$ animate\}. The results of the prior predictive simulations are shown in 
the lower part of Figure \ref{fig:model_predictions}. This model now shows the observed pattern: a semantic interference effect but no syntactic interference. One limitation here is that the prior predictive distribution of our post-hoc  model still overestimates the semantic interference effect (95\,\% CrI [10, 67]\,ms; the largest effect observed the SPR studywas [4, 12] ms at the critical region). This overestimation has to do with the fact that the prior on the F parameter is relatively spread out; it would be  possible to fit the parameter to the SPR data.

@@

%There is another point to make regarding the small size of the effects observed in the present study: The observed semantic effect in self-paced reading times (8 ms, 95\,\% CrI [4, 12]) only slightly overlaps with the quantitative predictions from the \citet{Lewis2005} model of cue-based retrieval for the \citet{vandyke07} design (95\,\% CrI [9, 49]\,ms, implementation of the model by \cite{engelmann_etal_2019}, adapted to the current design with the retrieval cues \{± animate\} and \{± grammatical subject\}). The \citet{Lewis2005} model does not only overestimate the effect size of semantic interference, but it also predicts equally sized semantic and syntactic interference effects (see Figure \ref{fig:model_predictions}). The syntactic effect is absent in our data.

@@

The comparison of the results of our SPR and ERP experiments suggests that the SPR results were strongly affected by encoding interference, while the ERPs elicited by the critical verb were not. This difference between methods might be caused by the difference in presentation rate. In the self-paced reading experiment, as implied by the name, the participants had control over the presentation rate of the stimuli. The observed slowdown starting at the distractor  might have resulted from the probably subconscious intent to slow down the sampling rate of the incoming stimuli to gain more time to process them. In contrast, in the ERP experiment, we used rapid serial visual presentation (RSVP), which presents stimuli at a fixed rate that is beyond the control of the participant. Therefore, they must adapt to the set rate and process the stimuli as fast as the rate requires. Here, it is noteworthy that the presentation duration of 500\,ms per word and 400\,ms between words which was used in the present ERP experiment is slower than usual natural reading speed. However, while in SPR the slowdown due to encoding and maintaining the distractor in memory overshadowed potential retrieval effects, the procedure of the ERP experiment might have prevented that. In the ERP results, we observed an effect at the critical verb, which could be attributed exclusively to the retrieval process hypothesized by cue-based retrieval accounts. Having said that, we cannot rule out the possibility that--due to the auto-paced nature of the ERP experiment--any encoding interference that started at or after the distractor is mixed in with the cost of retrieval interference. Indeed, recent modeling work on agreement attraction in reading \parencite{Yadavetal2022} has shown conclusively that some kind of feature distortion as well as cue-based retrieval are needed to explain the benchmark data-sets that exist so far on agreement attraction in reading. If the  \cite{Yadavetal2022} extends to interference effects in general, 
it is reasonable to assume that the effect observed at the verb in the ERP study is driven by both encoding and retrieval interference. However, we cannot resolve the relative roles of these latent processes in the present study.

@@

The evidence against a main effect of syntactic interference for subject-verb dependency resolution in our study is not consistent with the assumption in previous work, e.g., \cite{vandyke07,mertzen}, that the parser searches for a subject simply by setting the retrieval cue \{± grammatical subject\}.
  However, the semantic interference effect we found can be easily reconciled with previous work in  sentence processing. In the modeling section above, we have already discussed how our results relate to predictions from the \citet{Lewis2005} cue-based retrieval model.  
  %Our results may also be consisted with constraint-based models \citep[e.g.,][]{macdonald1994, mcrae1998modeling, trueswell1993}. Under constraint-based accounts, it is assumed that multiple constraints simultaneously and immediately influence processing. These accounts predict our ERP results: Compared to an inanimate distractor, an animate distractor -- regardless of its syntactic status -- would impede processing at a verb which requires an animate subject. The lack of a syntactic interference effect could indicate that the status `grammatical subject' was not used as a syntactic constraint. As discussed above, a plausible syntactic constraint is that a verb and its arguments need to be within the same clause. 
 Our results could also be seen as consistent with the good-enough processing account \citep{ferreira2007goodenough}, which assumes that comprehenders do not always aim to build a fully fleshed-out analysis of a sentence, but instead might accept incomplete, underspecified, or even incorrect analyses. The high number of participants that needed to be excluded in our experiments due to accuracy below 70\,\% (117 out of 908 participants in the SPR experiment, 29 out of 146 participants in the EEG experiment), suggests that our materials led to high processing demands. It is reasonable to assume that the participants might have--at least occasionally--adopted a good-enough processing mode when faced with high task demands \parencite{swets2008underspecification,LogacevMultiple,LogacevVasishthQJEP2016} or due to working memory capacity limitations \parencite{MalsburgVasishth2013}, or both. So, given a good-enough processing mode which leaves some syntactic relations within the sentence underspecified, it makes sense for the comprehender to use a simple heuristic for subject-verb dependency resolution. Animacy is an obvious choice for such a heuristic. Animate entities are proto-typical agents \citep{dowty1991thematic} and therefore, the primary use of the \{+ animate\} cue to retrieve a subject leads to the correct analysis with high probability (not just within the experiment context, but in everyday language use). Finally, our findings are also consistent with the possibility that language processing relies predominantly on semantic associations to form (probabilistic)
representation of event structures. This view has been put forward by \cite{rabovsky_etal_2018}, when they used the neural-network sentence gestalt model \parencite{mcclelland1989_sentence_gestalt}, to model the N400 amplitude. This model assumes that language comprehension relies on associative form-to-meaning mapping instead of syntactic rules. This assumption is consistent with our results.

Without the modeling results reported above, it would be easy to conclude that our data show that  syntactic cues play no role in sentence processing. However, this conclusion would be obviously wrong. First, as we showed above, the retrieval model suggests that--at least in our data--the subject cue is used, but differently than previously assumed: the relevant cue may be subject-in-same-clause. Second, there is plenty of independent evidence for the central role of syntactic information in comprehension; some examples are the role of case marking \parencite[e.g.,][]{ALV2019,HusainEtAl2014,bhatia2022processing,bader2000case,bader2006case,miyamoto2002case}, syntactic constraints like Principle A of the binding theory \parencite[e.g.,][]{Sturt2003,dillon2013,yadav2021individual},  and 
the importance of word order \parencite[e.g.,][]{meng2000mode}. There are of course important examples of the parser ignoring syntactic information. Three dramatic examples are the grammaticality illusion in multiple embeddings in English \parencite{gibsonthomas99}, the agreement attraction phenomenon across different languages \parencite{wagersetal,lago_etal_2021,jaeger_etal_2017}, and local coherence effects \parencite{taboretal04}.  However, there are languages (such as German and Dutch) in which the grammaticality illusion does not occur \parencite{VSLK11,FrankEtAl2015}, and a large-sample study \parencite{lcpaape2023} presents evidence against the syntactic local coherence effect that was reported by \cite{taboretal04}. Also, there seems to be at least one language that has rich case marking (Czech) that does not show any evidence of agreement attraction \parencite{chromy2023number}. Thus, it seems that, cross-linguistically, the role of syntax in building incremental structure and in completing dependencies remains as important as ever.

>>>>

% Useful packages
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
%\usepackage[colorlinks=true, allcolors=blue]{hyperref}
\usepackage[natbib=true,maxcitenames=2,bibstyle=authoryear, style=apa]{biblatex}
\usepackage{csquotes}
\DeclareLanguageMapping{american}{american-apa}
\bibliography{interferencebib}

@@

By contrast, syntactic interference is expected to be triggered if the distractor matches the \{+grammatical subject\} cue. This situation can occur when the relative clause  in \ref{ex:1_vandyke07} that modifies \textit{the resident} were to be changed to \textit{who said that the warehouse was dangerous}. Now, the distractor matches the \{+grammatical subject\} cue.

@@

Previous work has shown clear evidence for this kind of retrieval interference \parencite{vandyke07, vandyke_mcelree06, vandyke_mcelree2011, nicenboim,  vandyke_lewis03,jaeger_etal_2017}.
However, two important questions remain unanswered. First, as discussed in \textcite{jaeger_etal_2017}, almost all the published studies on interference effects are severely underpowered. Low power has the consequence that published estimates will tend to be overestimates or misestimates of the effect size \parencite{vasishth2018_signficancefilter}. Over-/misestimates have important implications for theory evaluation. One example is discussed in \textcite{vasishth2018_signficancefilter}, where a key prediction of surprisal theory \parencite{levy&keller_2013} could not be validated when a larger-sample study was conducted; by contrast, the original small-sample study  \parencite{levy&keller_2013} showed suggestive evidence consistent with the theory. Another example is the recent heated debate on prediction in sentence comprehension, which was largely focused on the results of low-powered studies, leading to potentially misleading conclusions \parencite{nicenboim_etal_2020}.
Second, although it is clear and uncontroversial  \parencite[e.g.,][]{vandyke07,mertzen} that syntactic and semantic cues are involved in sentence processing, it is not obvious which cues are relevant for a particular retrieval event. For example, \textcite{dillon2013} argued, following \textcite{Sturt2003}, that the gender cue was either downweighted or not used at all when an antecedent-reflexive dependency needs to be built; instead, the antecedent of a reflexive could be identified using the syntactic principle A of the binding theory, which correctly identifies the target for retrieval.  Using computational modeling, \textcite{dillon2013} suggested that antecedent-reflexive dependencies were immune to interference, and that the reason for this is that principle A of the binding theory is the primary driver of the cue-based retrieval process \parencite[cf. ][]{jaeger_etal_2020,yadav2021individual}. This was a surprising discovery because a priori, cue-based retrieval accounts do not allow for exceptions to retrieval interference in particular syntactic dependency types.
More broadly, an important implication of the work initiated by \textcite{dillon2013}, among others, is that  the precise details of which cues are used during retrieval can affect processing difficulty in possibly surprising ways.

@@

All experiments used the 2 $\times$ 2 design developed by \textcite{vandyke07}; the design crosses syntactic and semantic interference.

@@

Syntactic interference is manipulated via the syntactic status of the distractor, which is either subject (\ref{ex:hisyn}) of an embedded clause, or part of a prepositional phrase (\ref{ex:losyn}). Following \textcite{vandyke07}, the assumption here is that  during the retrieval of the subject, a distractor in any subject position causes more syntactic interference than a distractor inside a prepositional phrase.

@@

Indeed, previous work using the same design as in the present paper has also found semantic interference effects at the pre-critical region: both \textcite{vandyke07} and \textcite{mertzen} found such effects. \citeauthor{mertzen}'s (\citeyear{mertzen}) Figures 5 and 6 indicated reading time differences in earlier regions, especially for their German data (see their Figure 6). \textcite{vandyke07} did not report reading times for the whole sentences / distractors;  \citeauthor{vandyke07} attributed the effects observed at the pre-critical region to plausibility differences between conditions, but as \textcite{mertzen} also points out, encoding interference could be an explanation even in that study. Given these earlier findings, our results are consistent with the encoding interference explanation.

\textcite{mertzen} observed both syntactic and semantic interference effects in the pre-critical region. 
They discuss several alternative explanations for their observed effects. In addition to encoding interference, \textcite{mertzen} propose three other alternative explanations for their data: parafoveal-on-foveal effects, sentence structure confounds across conditions, and predictive processing effects. Our results cannot be explained by the parafoveal-on-foveal explanation, because there is no parafoveal preview in self-paced reading. The observed difference between the high and low semantic interference conditions is independent of the syntactic manipulation. Therefore, the sentence structure confound is also not a good explanation for the effect in our study. Regarding the predictive-processing explanation, \citet{mertzen} argued that the pre-critical adverb must attach to the upcoming the verbal phrase; this leads to an anticipatory creation of a verb phrase chunk in memory, which triggers a retrieval of the subject already at the pre-critical region. However, in our study, the effect started at the distractor and became smaller as the two pre-critical adverbs were read (see Figure \ref{fig:whole_sentence} A and B). The predictive-processing explanation would incorrectly predict that the effect begins at the second adverb, which immediately precedes the verb. Consequently, the prediction explanation is also ruled out in our data, rendering encoding interference as the most likely explanation of the effects found prior to the critical verb in our data. Crucially this line of reasoning should not be limited to the explanation of the pre-critical effects, but should be extended to the effects in the critical region as well. The reading times for high vs.\ low semantic interference conditions started to differ at the distractor and differed throughout the rest of the sentence. It is not possible to pinpoint whether the effect in the critical region is due to a retrieval initiated at this region, or still due to encoding and maintaining the distractor in memory, or a combination of the two kinds of interference.

\subsection{Comparison with previous findings}
The results of the present study stand in contrast to previous findings \citep{vandyke07, mertzen}. The present study is the only one to find semantic and no syntactic interference effect. The difference in observed effects could be explained by differences in statistical power or by methodological differences. The present study has higher statistical power than the previous studies. \citet{vandyke07} had 35 - 40 participants and 36 - 48 items in each of her three experiments, respectively. \textcite{mertzen} tested 61 English speakers and 121 German speakers with 40 items each. In contrast, we tested 774 participants with at least 60 items each (a subset of 160 participants read all 120 items). Therefore, the present study is the study with the highest power to date using this 2 $\times$ 2 interference design. Consequently, the estimates from the present study are the most precise ones so far for this design. The smaller effect size in the present study is a consequence of higher power, and the magnitude of the effects observed is comparable to that of other large-scale reading studies \citep{nicenboim} and meta-analyses on interference \citep{jaeger_etal_2017}.

@@

Models were run with four chains and 20,000 iterations of which 2,000 were used as warm-up phase \citep{schad_etal_2022_BF}. For the calculation of Bayes factors and the corresponding sensitivity analysis, we defined a range of priors on the parameters of interest, assuming a range of effect sizes \citep{nicenboim_stats}. We chose slope priors which assume a wide range of effect sizes from $-0.2$ to $-11$\,$\mu V$. However, the more realistic a priori effect sizes probably range from $-0.2$ to at most $-2$\,$\mu V$; this is based on our expectation that the interference effect investigated here, being a subtle manipulation, is probably smaller in magnitude than the well-established  effect of cloze probability on the N400 amplitude analyzed by \textcite{nicenboim_stats}, which ranges from $0$ to $5$\,$\mu V$ \parencite[see section 5.2.5][]{nicenboim_stats}.

@@

Consistent with the self-paced reading study, there is clear evidence for semantic interference, and evidence against syntactic interference.  How do these observed patterns compare to the theoretical predictions of cue-based retrieval theory? We turn to this point next by discussing the quantiative predictions of the \textcite{Lewis2005} model.

%The to date rather small body of ERP studies on interference effects during sentence comprehension gives an indeterminate picture of the nature of interference effects in ERP measures. The present study adds a large data point to the body of literature which reported negativities or more specifically N400 effects for high vs.\ low interference conditions \citep{vasishth_drenhaus_2011, martinetal2014, schoknecht2022, lee_garnsey}. Therefore, it contributes to the debate of how interference effects manifest themselves electrophysiologically (interference effects have also been reported in differences in the P600, \textcite{Tanner_etal_2017, martinetal2014}).

@@

 \textcite{mertzen} presented simulations that show approximately equally sized interference effects 
for both the syntactic and semantic manipulations. Their model assumed that three cues were used at the verb:  \{$\pm$ subject, $\pm$ same clause, $\pm$ animacy\}. The upper part of Figure \ref{fig:model_predictions} reproduces the \textcite{mertzen}   simulations. These are prior predictive distributions of the effects of interest, by sampling from a prior distribution ($Beta(a=4,b=6)$) for the free scaling parameter F and then generating data from the model while holding all other parameters in the model at fixed values. The F parameter scales the activation to a retrieval time. The prior $Beta(a=4,b=6)$ was chosen in earlier work \parencite{jaeger_etal_2020}.

@@

The absence of (and evidence against) syntactic interference is surprising because the psycholinguistic literature has clearly established that syntactic constraints play a role in sentence processing. For example, \textcite{dillon2013} and \textcite{Sturt2003} argued that principle A of the binding theory controls the retrieval process in the antecedent-reflexive dependency, effectively eliminating or attenuating interference from other overlapping features (such as number marking) of intervening nouns that cannot be an antecedent for the reflexive. For further evidence for this idea from individual differences, see \textcite{yadav2021individual}. Even more relevant to the present study is the fact that \textcite{mertzen} did find syntactic interference effects.

Could the absence of syntactic interference be reconciled with a cue-based retrieval account? We speculated that one possibility might be that the parser tracks hierarchical structure and that this hierarchical information is used when searching for a noun in memory. Specifically, the relevant syntactic cues that might be used at the verb is not  \{$\pm$ subject, $\pm$ same clause\} but rather a composite cue  \{$\pm$ subject-in-same-clause\} . In other words, in the spirit of the  \textcite{dillon2013} and \textcite{Sturt2003} proposal, that the parser intelligently targets only the syntactically relevant noun during antecedent retrieval, we evaluated the predictions of the model assuming that the cues are   \{$\pm$ subject-in-same-clause, $\pm$ animate\}. The results of the prior predictive simulations are shown in 
the lower part of Figure \ref{fig:model_predictions}. This model now shows the observed pattern: a semantic interference effect but no syntactic interference. One limitation here is that the prior predictive distribution of our post-hoc  model still overestimates the semantic interference effect (95\,\% CrI [10, 67]\,ms; the largest effect observed the SPR studywas [4, 12] ms at the critical region). This overestimation has to do with the fact that the prior on the F parameter is relatively spread out; it would be  possible to fit the parameter to the SPR data.

@@

%There is another point to make regarding the small size of the effects observed in the present study: The observed semantic effect in self-paced reading times (8 ms, 95\,\% CrI [4, 12]) only slightly overlaps with the quantitative predictions from the \citet{Lewis2005} model of cue-based retrieval for the \citet{vandyke07} design (95\,\% CrI [9, 49]\,ms, implementation of the model by \textcite{engelmann_etal_2019}, adapted to the current design with the retrieval cues \{± animate\} and \{± grammatical subject\}). The \citet{Lewis2005} model does not only overestimate the effect size of semantic interference, but it also predicts equally sized semantic and syntactic interference effects (see Figure \ref{fig:model_predictions}). The syntactic effect is absent in our data.

@@

The comparison of the results of our SPR and ERP experiments suggests that the SPR results were strongly affected by encoding interference, while the ERPs elicited by the critical verb were not. This difference between methods might be caused by the difference in presentation rate. In the self-paced reading experiment, as implied by the name, the participants had control over the presentation rate of the stimuli. The observed slowdown starting at the distractor  might have resulted from the probably subconscious intent to slow down the sampling rate of the incoming stimuli to gain more time to process them. In contrast, in the ERP experiment, we used rapid serial visual presentation (RSVP), which presents stimuli at a fixed rate that is beyond the control of the participant. Therefore, they must adapt to the set rate and process the stimuli as fast as the rate requires. Here, it is noteworthy that the presentation duration of 500\,ms per word and 400\,ms between words which was used in the present ERP experiment is slower than usual natural reading speed. However, while in SPR the slowdown due to encoding and maintaining the distractor in memory overshadowed potential retrieval effects, the procedure of the ERP experiment might have prevented that. In the ERP results, we observed an effect at the critical verb, which could be attributed exclusively to the retrieval process hypothesized by cue-based retrieval accounts. Having said that, we cannot rule out the possibility that--due to the auto-paced nature of the ERP experiment--any encoding interference that started at or after the distractor is mixed in with the cost of retrieval interference. Indeed, recent modeling work on agreement attraction in reading \parencite{Yadavetal2022} has shown conclusively that some kind of feature distortion as well as cue-based retrieval are needed to explain the benchmark data-sets that exist so far on agreement attraction in reading. If the  \textcite{Yadavetal2022} extends to interference effects in general, 
it is reasonable to assume that the effect observed at the verb in the ERP study is driven by both encoding and retrieval interference. However, we cannot resolve the relative roles of these latent processes in the present study.

@@

The evidence against a main effect of syntactic interference for subject-verb dependency resolution in our study is not consistent with the assumption in previous work, e.g., \textcite{vandyke07,mertzen}, that the parser searches for a subject simply by setting the retrieval cue \{± grammatical subject\}.
  However, the semantic interference effect we found can be easily reconciled with previous work in  sentence processing. In the modeling section above, we have already discussed how our results relate to predictions from the \citet{Lewis2005} cue-based retrieval model.  
  %Our results may also be consisted with constraint-based models \citep[e.g.,][]{macdonald1994, mcrae1998modeling, trueswell1993}. Under constraint-based accounts, it is assumed that multiple constraints simultaneously and immediately influence processing. These accounts predict our ERP results: Compared to an inanimate distractor, an animate distractor -- regardless of its syntactic status -- would impede processing at a verb which requires an animate subject. The lack of a syntactic interference effect could indicate that the status `grammatical subject' was not used as a syntactic constraint. As discussed above, a plausible syntactic constraint is that a verb and its arguments need to be within the same clause. 
 Our results could also be seen as consistent with the good-enough processing account \citep{ferreira2007goodenough}, which assumes that comprehenders do not always aim to build a fully fleshed-out analysis of a sentence, but instead might accept incomplete, underspecified, or even incorrect analyses. The high number of participants that needed to be excluded in our experiments due to accuracy below 70\,\% (117 out of 908 participants in the SPR experiment, 29 out of 146 participants in the EEG experiment), suggests that our materials led to high processing demands. It is reasonable to assume that the participants might have--at least occasionally--adopted a good-enough processing mode when faced with high task demands \parencite{swets2008underspecification,LogacevMultiple,LogacevVasishthQJEP2016} or due to working memory capacity limitations \parencite{MalsburgVasishth2013}, or both. So, given a good-enough processing mode which leaves some syntactic relations within the sentence underspecified, it makes sense for the comprehender to use a simple heuristic for subject-verb dependency resolution. Animacy is an obvious choice for such a heuristic. Animate entities are proto-typical agents \citep{dowty1991thematic} and therefore, the primary use of the \{+ animate\} cue to retrieve a subject leads to the correct analysis with high probability (not just within the experiment context, but in everyday language use). Finally, our findings are also consistent with the possibility that language processing relies predominantly on semantic associations to form (probabilistic)
representation of event structures. This view has been put forward by \textcite{rabovsky_etal_2018}, when they used the neural-network sentence gestalt model \parencite{mcclelland1989_sentence_gestalt}, to model the N400 amplitude. This model assumes that language comprehension relies on associative form-to-meaning mapping instead of syntactic rules. This assumption is consistent with our results.

Without the modeling results reported above, it would be easy to conclude that our data show that  syntactic cues play no role in sentence processing. However, this conclusion would be obviously wrong. First, as we showed above, the retrieval model suggests that--at least in our data--the subject cue is used, but differently than previously assumed: the relevant cue may be subject-in-same-clause. Second, there is plenty of independent evidence for the central role of syntactic information in comprehension; some examples are the role of case marking \parencite[e.g.,][]{ALV2019,HusainEtAl2014,bhatia2022processing,bader2000case,bader2006case,miyamoto2002case}, syntactic constraints like Principle A of the binding theory \parencite[e.g.,][]{Sturt2003,dillon2013,yadav2021individual},  and 
the importance of word order \parencite[e.g.,][]{meng2000mode}. There are of course important examples of the parser ignoring syntactic information. Three dramatic examples are the grammaticality illusion in multiple embeddings in English \parencite{gibsonthomas99}, the agreement attraction phenomenon across different languages \parencite{wagersetal,lago_etal_2021,jaeger_etal_2017}, and local coherence effects \parencite{taboretal04}.  However, there are languages (such as German and Dutch) in which the grammaticality illusion does not occur \parencite{VSLK11,FrankEtAl2015}, and a large-sample study \parencite{lcpaape2023} presents evidence against the syntactic local coherence effect that was reported by \textcite{taboretal04}. Also, there seems to be at least one language that has rich case marking (Czech) that does not show any evidence of agreement attraction \parencite{chromy2023number}. Thus, it seems that, cross-linguistically, the role of syntax in building incremental structure and in completing dependencies remains as important as ever.
